Density-dependent selection and the limits of relative fitness

### 2018-02-26 manuscript.tex
l

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. Suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories contribute negligibly to each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, these identical types should retain constant relative frequencies, and so the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.

This consistency requirement is not automatically satisfied even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually good approximations. Even small violations will cause one of the identical types to grow exponentially relative to the other. This pathological growth occurs purely because one group is treated as rare and the other as common. Since a single-type population can be divided into arbitrary rare and common subsets of identical types, and clearly not all such subsets can grow/decline, we have a logical contradiction.

>>>>

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types. To illustrate, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$. Since $L$ is large, uncontested territories contribute negligibly to each type's growth. The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal. Even small violations of this consistency condition will cause one type to grow exponentially relative to the other. This is a logical contradiction: any single-type population can be divided at will into rare and common subsets of identical types, and clearly not all such subsets can grow/decline.
